---
title: "ST 558 Project 1"
author: Ryan Clark
format: html
---

## Intro

In this project we will query the Public Use Microdata Sample (PUMS) Census API. This API provides psuedo person-level data from the American Community Survey. We will focus on retrieving and summarizing data from four survey years: 2021, 2022, 2023 and 2024. 

## Part 1: Data Processing

Every year surveyed by the Census API has its own `variables.json` endpoint that lays out each variable used and their respective definitions. While there is plenty of overlap year-to-year, important differences exist that we must account for when parsing variable values from the JSON files. In this first step, we build a function that retrieves these files, extracts only the variables relevant to our analysis, and stores the result so we can reference it later.

```{r}
# Function to `GET` a PUMS `variables.json` endpoint 
get_pums_variables <- function(url, vars_to_keep) {
  var_info <- httr::GET(url)  # Contact API 
  parsed_vars <- jsonlite::fromJSON(rawToChar(var_info$content))  # Parse the raw JSON
  parsed_vars$variables[vars_to_keep]
}

# Variables found in all four years
common_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
                  "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
                  "REGION", "DIVISION")

# ST/STATE variable changes by year, so append to each year's variable list
vars_2021 <- c(common_vars, "ST")
vars_2022 <- c(common_vars, "ST")
vars_2023 <- c(common_vars, "STATE")
vars_2024 <- c(common_vars, "STATE")

# URLs
url_2021 <- "https://api.census.gov/data/2021/acs/acs1/pums/variables.json"
url_2022 <- "https://api.census.gov/data/2022/acs/acs1/pums/variables.json"
url_2023 <- "https://api.census.gov/data/2023/acs/acs1/pums/variables.json"
url_2024 <- "https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

# Build one list per year, then merge into a single list
pums_variables <- list(
  "2021" = get_pums_variables(url_2021, vars_2021),
  "2022" = get_pums_variables(url_2022, vars_2022),
  "2023" = get_pums_variables(url_2023, vars_2023),
  "2024" = get_pums_variables(url_2024, vars_2024)
)

# Save list as an .rds in the `data` folder
saveRDS(pums_variables, file = "data/pums_variables.rds")
```


```{r}
#| include: false
# API key
census_key <- Sys.getenv("CENSUS_API_KEY")
```

(placeholder text)

```{r}
# TESTING LOGIC AND STRUCTURE
# Function to query the PUMS API, return a parsed tibble. 
query_pums_data <- function(year = 2024,
                             numeric_vars = "AGEP",
                             categorical_vars = "SEX",
                             geography = "State",
                             geo_subset = NULL) {

  # Validate year
  valid_years <- c(2021, 2022, 2023, 2024)
  if (length(year) != 1 || !(year %in% valid_years)) {  # Not exactly 1 year OR not in list
    stop("Error: `year` must be 2021, 2022, 2023, or 2024.")
  }

  # Validate numeric variables
  valid_numeric <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if (length(numeric_vars) < 1) {  # Empty input
    stop("Error: At least one numeric variable (other than PWGTP) must be specified.")
  }
  if (!all(numeric_vars %in% valid_numeric)) {  # Checking if any vars are invalid
    stop("Error: Invalid numeric variable(s) provided. Valid options are: AGEP, GASP, GRPIP, JWAP, JWDP, and JWMNP.")
  }
  # PWGTP always returned
  numeric_vars <- unique(c(numeric_vars, "PWGTP"))

  # Validate categorical variables
  valid_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if (length(categorical_vars) < 1) {  # Empty input
    stop("Error: At least one categorical variable must be specified.")
  }
  if (!all(categorical_vars %in% valid_categorical)) {  # Checking if any vars are invalid
    stop("Error: Invalid categorical variable(s) provided. Valide options are: FER, HHL, SCH, SCHL, and SEX.")
  }

  # Validate geography level
  valid_geo <- c("Region", "Division", "State")
  if (length(geography) != 1 || !(geography %in% valid_geo)) {  # Not exactly 1 level OR not in list
    stop("Error: `geography` must be exactly Region, Division, or State.")
  }

  # Geo_subset default
  default_geo_subset <- list(Region = "3", Division = "5", State = "37")
  if (is.null(geo_subset)) {  # Empty input
    geo_subset <- default_geo_subset[[geography]] # Picks the correct default value based on the chosen geo
  }
  # NOTE: validate geo_subset against .rds file here?

  # Return inputs to verify
  list(
    year = year,
    numeric_vars = numeric_vars,
    categorical_vars = categorical_vars,
    geography = geography,
    geo_subset = geo_subset
  )
}
```










