---
title: "ST 558 Project 1"
author: Ryan Clark
date: 2026-09-30
abstract: |
  The purpose of this report is to write R functions that query the PUMS Census API and summarize and plot the returned data.
format: html
---

# Intro

In this project we will query the Public Use Microdata Sample (PUMS) Census API. This API provides pseudo person-level data from the American Community Survey. We will focus on retrieving, summarizing, and plotting data from four survey years: 2021, 2022, 2023 and 2024. In Part 1 we write functions to query the API and return tibbles. In Part 2 we write custom `summary()` and `plot()` functions to generate results.

# Part 1: Data Processing

## Parsing Variable Values

Each survey year from the Census API has its own `variables.json` endpoint that lays out each variable used and their respective definitions. While there is plenty of overlap year-to-year, important differences exist that we must account for when parsing variable values from the JSON files. In this first step, we build a function that retrieves these files, extracts only the variables relevant to our analysis, and stores the result so we can reference it later.

```{r}
# Function to `GET` a PUMS `variables.json` endpoint 
get_pums_variables <- function(url, vars_to_keep) {
  var_info <- httr::GET(url)  # Contact API 
  parsed_vars <- jsonlite::fromJSON(rawToChar(var_info$content))  # Parse the raw JSON
  parsed_vars$variables[vars_to_keep]
}

# Variables found in all four years
common_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
                  "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
                  "REGION", "DIVISION")

# ST/STATE variable changes by year, so append to each year's variable list
vars_2021 <- c(common_vars, "ST")
vars_2022 <- c(common_vars, "ST")
vars_2023 <- c(common_vars, "STATE")
vars_2024 <- c(common_vars, "STATE")

# URLs
url_2021 <- "https://api.census.gov/data/2021/acs/acs1/pums/variables.json"
url_2022 <- "https://api.census.gov/data/2022/acs/acs1/pums/variables.json"
url_2023 <- "https://api.census.gov/data/2023/acs/acs1/pums/variables.json"
url_2024 <- "https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

# Build one list per year, then merge into a single list
pums_variables <- list(
  "2021" = get_pums_variables(url_2021, vars_2021),
  "2022" = get_pums_variables(url_2022, vars_2022),
  "2023" = get_pums_variables(url_2023, vars_2023),
  "2024" = get_pums_variables(url_2024, vars_2024)
)

# Save list as an .rds in the `data` folder
saveRDS(pums_variables, file = "data/pums_variables.rds")
```


```{r}
#| include: false
# API key
census_key <- Sys.getenv("CENSUS_API_KEY")
```

## Parsing API Data

With our variable dictionaries saved, we can move on to querying the actual data. Before writing the main querying function, we introduce two helper functions to handle the conversions it will need. 

The first helper handles the time variables `JWAP` (time of arrival at work) and `JWDP` (time of departure for work). The API returns these as codes for time ranges, not actual clock time values. Therefore, our helper looks up each code's label, divides it into a start and an end time string, converts both to time values with `lubridate`, and returns the midpoint of the range. Codes whose label isn't a time range (e.g. "N/A" code) are returned as `NA`.

Importantly, the dictionary codes for these two variables have a fixed-width of three characters but the API drops leading zeros so we must pad them where applicable (e.g. `"7"` to `"007"`). The exception is the "N/A" code which is simply `"0"`.

```{r}
# Helper function: Handling the time-coded variables (JWAP/JWDP)
parse_pums_time <- function(codes, code_labels) {
  padded_codes <- ifelse(codes == "0", "0",  # If 0, keep (it's the N/A code)
                         formatC(as.integer(codes), width = 3, flag = "0"))  # Otherwise convert int and insert 0s to get fixed-width

  labels <- code_labels[padded_codes]  # Match padded code to dictionary label

  # Logical vector using language ("to") to identify time-range observations
  has_range <- !is.na(labels) & grepl(" to ", labels, fixed = TRUE)

  # Isolate times
  starts <- trimws(gsub("\\.", "", sub(" to .*$", "", labels[has_range])))  # Before "to"
  ends   <- trimws(gsub("\\.", "", sub("^.*? to ", "", labels[has_range])))  # After "to"

  # Convert string times to datetime objects
  start_times <- lubridate::parse_date_time(starts, orders = "IMp", tz = "UTC")  
  end_times   <- lubridate::parse_date_time(ends,   orders = "IMp", tz = "UTC")

  # Compute midpoint time
  mid_times <- start_times + (end_times - start_times) / 2

  # Initialize empty datetime vector
  result <- as.POSIXct(rep(NA, length(labels)), tz = "UTC")
  # Fill it in with midpoint values
  result[has_range] <- mid_times
  # Return vector
  result
}
```

The second helper handles our categorical variables. Like the time variables above, the categorical variables are delivered as raw codes from the API, so the helper replaces them with their corresponding labels (accounting for year-to-year differences). It then converts the result into a factor whose levels mimic the order of the original codes. 

These codes have the same leading-zero problem as the time variables, though without a fixed-width and not quite as pervasive. So we must manually grab the proper width each time before padding. This function will be called for every categorical variable requested by the user.

```{r}
# Helper function: Replaces categorical raw character codes with corresponding label
pums_to_factor <- function(codes, code_labels) {
  # Width used by this variable's dictionary keys, excluding "0" (N/A)
  pad_width <- max(nchar(setdiff(names(code_labels), "0")))

  # Add leading zeros; except "0" (N/A) is never padded
  padded_codes <- ifelse(codes == "0", "0", formatC(as.integer(codes), width = pad_width, flag = "0"))
  
  ordered_codes  <- names(code_labels)[order(as.numeric(names(code_labels)))]  # Numerical ordering of codes
  ordered_labels <- unname(code_labels[ordered_codes])  # Labels ordered by their codes
  
  # Return categorical factor
  factor(unname(code_labels[padded_codes]), levels = ordered_labels)
}
```

Now that both helpers have been established, we can write the main query function that will utilize them. This function lets the user choose what variables to request from the API, ensures these are valid choices, and returns a tibble. Note that we are only parsing a specific subset of the entire list of variables (see code below). If the inputs pass all checks, then the API URL is built and our API key makes the request. From there, the function converts the JSON into a cleaned tibble with a `"census"` class.

```{r}
# Function to query the PUMS API (year, variables, and geography) then return a parsed tibble. 
query_pums_data <- function(year = 2024,
                            numeric_vars = "AGEP",
                            categorical_vars = "SEX",
                            geography = "State",
                            geo_subset = NULL) {

  # Validate year
  valid_years <- c(2021, 2022, 2023, 2024)
  if (length(year) != 1 || !(year %in% valid_years)) {  # Not exactly 1 year OR not in list
    stop("Error: `year` must be 2021, 2022, 2023, or 2024.")
  }

  # Validate numeric variables
  valid_numeric <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if (length(numeric_vars) < 1) {  # Empty input
    stop("Error: At least one numeric variable (other than PWGTP) must be specified.")
  }
  if (!all(numeric_vars %in% valid_numeric)) {  # Checking if any vars are invalid
    stop("Error: Invalid numeric variable(s) provided. Valid options are: AGEP, GASP, GRPIP, JWAP, JWDP, and JWMNP.")
  }
  # PWGTP always returned
  numeric_vars <- unique(c(numeric_vars, "PWGTP"))

  # Validate categorical variables
  valid_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if (length(categorical_vars) < 1) {  # Empty input
    stop("Error: At least one categorical variable must be specified.")
  }
  if (!all(categorical_vars %in% valid_categorical)) {  # Checking if any vars are invalid
    stop("Error: Invalid categorical variable(s) provided. Valide options are: FER, HHL, SCH, SCHL, and SEX.")
  }

  # Validate geography level
  valid_geo <- c("Region", "Division", "State")
  if (length(geography) != 1 || !(geography %in% valid_geo)) {  # Not exactly 1 level OR not in list
    stop("Error: `geography` must be exactly Region, Division, or State.")
  }

  # Geo_subset default
  default_geo_subset <- list(Region = "3", Division = "5", State = "37")  # North Carolina
  if (is.null(geo_subset)) {  # Empty input
    geo_subset <- default_geo_subset[[geography]] # Picks the correct default value based on the chosen geo
  }

  geo_for_name <- tolower(geography)  # Convert to lowercase for URL syntax
  geo_var_name <- switch(geography,  # Col name to return
    "Region" = "REGION",
    "Division" = "DIVISION",
    "State" = if (year %in% c(2021, 2022)) "ST" else "STATE"
  )

  # String for the URL `for=`
  geo_value <- if (identical(geo_subset, "all")) "*"  # All values (API syntax)
               else geo_subset  # User's specific choice

  # Variables list
  get_vars <- unique(c(numeric_vars, categorical_vars, geo_var_name))

  # URL
  base_url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums")
  url <- paste0(
    base_url,
    "?get=", paste(get_vars, collapse = ","),
    "&for=", geo_for_name, ":", geo_value,
    "&key=", census_key
  )

  # Contact API endpoint
  resp <- httr::GET(url)
  raw_data <- jsonlite::fromJSON(rawToChar(resp$content))

  # Initialize tibble
  pums_tibble <- raw_data |>
  tibble::as_tibble(.name_repair = "minimal") |>  # Prevent auto-naming
  janitor::row_to_names(row_number = 1)  # First row is col headers, removed from data
  pums_tibble[[geo_for_name]] <- NULL  # Drop the geography duplicate column
  
  # Load RDS and convert year to char
  variable_info <- readRDS("data/pums_variables.rds")
  year_info <- variable_info[[as.character(year)]]

  # Convert numeric variables to num or POSIXct
  time_vars <- c("JWAP", "JWDP")  # Arrival/departure times
  for (var in numeric_vars) {
    if (var %in% time_vars) {  # is the variable time-coded?
      code_labels <- unlist(year_info[[var]]$values$item)  # Table of codes and labels
      pums_tibble[[var]] <- parse_pums_time(pums_tibble[[var]], code_labels)  # Helper function
    } else {
      pums_tibble[[var]] <- as.numeric(pums_tibble[[var]]) # Otherwise numeric
    }
  }

  # Convert categorical variables to labeled factors
  for (var in categorical_vars) {
    code_labels <- unlist(year_info[[var]]$values$item)  # Table of codes and labels
    pums_tibble[[var]] <- pums_to_factor(pums_tibble[[var]], code_labels) # Helper function
  }

  # Adding the "census" class to the tibble
  class(pums_tibble) <- c("census", class(pums_tibble))
  
  # Return tibble
  pums_tibble
}
```

Let's show some examples of how this querying process might look for a user.

This example will not use default variables. We will use `GASP` (gas cost) and `JWMNP` (travel time to work) as numeric variables, `SCHL` (educational attainment) as a categorical variable, and the Pacific census division (division code `"9"`) for the 2023 survey year.

```{r}
# Example Query 
example_single_year <- query_pums_data(
  year = 2023,
  numeric_vars = c("GASP", "JWMNP"),
  categorical_vars = "SCHL",
  geography = "Division",
  geo_subset = "9"
)
example_single_year
```

See that `PWGTP` is printed even though we only specified `GASP` and `JWMNP`. This is because we append `PWGTP` to the `numeric_vars` list so it's always returned. `SCHL` is a factor with its levels pulled from the 2023 variable dictionary instead of raw numeric codes, and the `DIVISION` column reflects the geography we subset on.

We also setup our function to validate the user inputs and print an error message when appropriate. For example, requesting a survey year outside 2021–2024 yields the following:

```{r, error = TRUE}
query_pums_data(year = 2020)
```

## Combining Years

Notice that the query function above only accepts one year as an argument. Clearly this is not ideal, so we expand on this by building another query function below that can take a vector of years. This new query still relies on the previous function to go through each year's call. Our query below accounts for variable differences between years, makes a `year` column identifying the source year for a given row, and then stacks the results into a tibble while ultimately preserving the `"census"` attribute.

```{r}
# Function to query multiple data years then return a tibble.
query_pums_multi_year <- function(years,  # Vector of years (unlike query_pums_data)
                                  numeric_vars = "AGEP",
                                  categorical_vars = "SEX",
                                  geography = "State",
                                  geo_subset = NULL) {

  # Loop over each element of `years`, applies function to assemble list of tibbles
  results <- lapply(years, function(yr) {
    result <- query_pums_data(  # A year's raw tibble
      year = yr,
      numeric_vars = numeric_vars,
      categorical_vars = categorical_vars,
      geography = geography,
      geo_subset = geo_subset
    )

    # Changing state variable names to all be "STATE" (2021/2022)
    if ("ST" %in% names(result)) {
      names(result)[names(result) == "ST"] <- "STATE"
    }

    # Record which survey year this data came from
    result$year <- yr

    # Strip the "census" class before combining (for dplyr)
    class(result) <- class(result)[class(result) != "census"]
    
    # Return a year's tibble
    result
  })
  
  # Combine (vertically)
  combined <- dplyr::bind_rows(results)

  # Adding back the "census" class to the combined result
  class(combined) <- c("census", class(combined))

  # Return tibble
  combined
}
```

Let's look at an example of combining several years (2021-2023) into one tibble. We can use this as an opportunity to check the state variable change (`ST` to `STATE`).

```{r}
# Example Query
example_multi_year <- query_pums_multi_year(
  years = c(2021, 2022, 2023),
  numeric_vars = "AGEP",
  categorical_vars = "SEX"
)
example_multi_year
```

We have a `year` column to identify which survey year each observation came from. The geography column is consistently named `STATE` throughout, which is what we intended. Without renaming `ST` to `STATE` we would have two geography columns (`ST` and `STATE`).

# Part 2: Writing a Generic Function for Summarizing

We've now established that we can query and combine PUMS data, so next we attempt to summarize this data and eventually visualize it. To make our summaries convenient to use, we will take advantage of the `"census"` class we've been defining our tibbles with. I will elaborate on more on this below. 

## Generic Functions

In R there are functions that use a class-specific method to operate on their arguments. That is, how the function handles the argument provided is based on the class of said argument. These are known as *generic functions*. For example, `summary()` on a numeric vector and `summary()` on a linear model (`lm` object) behave entirely different because they're running different underlying functions — `summary.default()` versus `summary.lm()`. `Plot()` is another such example.

Recall in Part 1, every tibble returned by `query_pums_data()` and `query_pums_multi_year()` was tagged with `class(x) <- c("census", class(x))`. Therefore, if we write functions named `summary.census()` and `plot.census()`, R will automatically call these when a user runs `summary()` or `plot()` on a query result. This should be more efficient for generating the desired results.

## Numeric Summaries

We now construct our `summary.census()` function. This function will produce weighted means and standard deviations for numeric variables, excluding the time variables, and a count of rows at each level of each categorical variable. `PWGTP` is the weight and informs us how many people each row represents. Since we want to return a list, I decided to go with a two-part list: a mean/SD variable and a categorical variable. The user can choose which variables to summarize.

```{r}
# Census summary function
summary.census <- function(object, numeric_vars = NULL, categorical_vars = NULL, ...) {
  # Tibble col names
  all_names <- names(object)

  # Grabbing desired numeric vars (no weight or year)
  eligible_numeric <- setdiff(all_names[sapply(object, is.numeric)], c("PWGTP", "year"))

  # Grabbing categorical vars (factor columns, plus year if present)
  eligible_categorical <- all_names[sapply(object, is.factor)]
  if ("year" %in% all_names) {
    eligible_categorical <- c(eligible_categorical, "year")  # Appends year by name
  }

  # Default to above eligible variables if the user didn't specify
  if (is.null(numeric_vars)) numeric_vars <- eligible_numeric
  if (is.null(categorical_vars)) categorical_vars <- eligible_categorical

  # Validate variable names inputted by user
  if (!all(numeric_vars %in% all_names)) {
    stop("Error: `numeric_vars` not found.")
  }
  if (!all(categorical_vars %in% all_names)) {
    stop("Error: `categorical_vars` not found.")
  }

  weights <- object$PWGTP  # Survey weight

  # Weighted mean/sd for each numeric variable
  numeric_summary <- lapply(numeric_vars, function(var) {  # Applies function across numeric_vars and collects results into a list
    values <- object[[var]]  # Get each variable col vector
    # Mean
    weighted_mean <- sum(values * weights) / sum(weights)
    # SD
    weighted_sd <- sqrt(sum(values^2 * weights) / sum(weights) - weighted_mean^2)
    # Return vector
    c(mean = weighted_mean, sd = weighted_sd)
  })
  # Attaches variable names to the numeric list
  names(numeric_summary) <- numeric_vars

  # Counts for each categorical variable
  categorical_summary <- lapply(categorical_vars, function(var) {
    table(object[[var]])
  })
  # Attaches variable names to the categorical list
  names(categorical_summary) <- categorical_vars

  # Combines the two lists into one and returns it
  list(numeric = numeric_summary, categorical = categorical_summary)
}
```

We shall now test some examples while building off of the previous examples from Part 1. Recall that `example_single_year` has the `"census"` class, so calling the generic `summary()` function on it should implement our `summary.census()`. 

```{r}
# Example summary
summary(example_single_year)
```

Clearly from the mean/sd and count output, `summary()` called our `summary.census()` function. With no additional arguments, it defaults to every eligible variable: `GASP` and `JWMNP` get weighted means and standard deviations, and `SCHL` gets counted by level. `DIVISION` is not one of the categorical variables `summary.census()` recognizes so it correctly gets skipped.

Let's return to `example_multi_year` to check whether `year` is treated as categorical, as intended.

```{r}
# Example summary
summary(example_multi_year)
```

We can see that results from the years 2021-2023 are displayed as they should. And `year` appears as its own categorical breakdown alongside `SEX`. While `year` is stored as a plain numeric column, it's counted rather than averaged.

## Graphical Summaries

As we mentioned above, `plot()` is another generic function we can use for data visualizations. So we can write a `plot.census()` function for our `"census"` tibbles. The user will input one categorical variable and one numerical variable. With that, we will generate a boxplot of the numeric variable at the levels of the categorical variable. As with the numeric summaries, the rows are weighted by `PWGTP`, which means the plots will reflect the population

```{r}
library(ggplot2)

# Census plot function
plot.census <- function(x, categorical_var, numeric_var, ...) {

  # Validate variable names inputted by user
  if (!(categorical_var %in% names(x))) {
    stop("Error: `categorical_var` not found.")
  }
  if (!(numeric_var %in% names(x))) {
    stop("Error: `numeric_var` not found.")
  }

  # Plot
  ggplot(x, aes(x = as.factor(get(categorical_var)),  # When x=year we need it to be factor for plotting
                y = get(numeric_var), 
                weight = PWGTP)) +
    geom_boxplot() +
    labs(x = categorical_var, y = numeric_var)
}
```

As we saw with `summary()`, calling `plot()` on `example_multi_year` dispatches to `plot.census()` automatically. For this next example we'll continue to use `example_multi_year`, plotting `year` as the categorical variable and `AGEP` (age) as the numeric variable. 

```{r}
# Plot example
plot(example_multi_year, categorical_var = "year", numeric_var = "AGEP")
```

By default, `plot()` would've drawn scatterplots but since the argument was a `"census"` type it instead used our `plot.census()` to draw weighted boxplots. While these boxplots appear identical, they are built from different datasets. We can confirm this with the code below.

```{r}
# Checking the data that builds the previous example
example_multi_year |>
  dplyr::group_by(year) |>
  dplyr::summarize(
    rows = dplyr::n(),
    population = sum(PWGTP),
    weighted_mean_age = sum(AGEP * PWGTP) / sum(PWGTP),
    max_age = max(AGEP)
  )
```

We can see that these boxplots are built from different data weighted by `PWGTP`. We might explain the similarities by suggesting that year-to-year age distributions in one state (in this case North Carolina) are relatively stable. 

Let's do one more plot example, still building off of the previous multi-year query. This time we'll change the categorical variable to `SEX`, and we'll try sprucing up the plot a bit. We should be able to change the aesthetics because `plot.census()` returns a ggplot object that we can add layers to without modifying the function.

```{r}
# Plot example
plot(example_multi_year, categorical_var = "SEX", numeric_var = "AGEP") +
  aes(fill = SEX) +
  guides(fill = "none") +
  coord_flip()
```

Once again, we can infer that our `plot.census()` function was utilized based on the output. Combining all three surveys (2021-2023), the age distributions for males and females in North Carolina are quite similar. Women have a slightly older median and quartiles, although the whiskers are approximately equal between the sexes. It is a known fact that women live longer than men, which may be driving these differences. This plot limits what we can see at the older ages because the whiskers simply display the highest age recorded for both sexes.

# Conclusion

In this project we built a set of functions that allow the user to query the Public Use Microdata Sample Census API between the years 2021-2024, select which data they want, then summarize and plot this data. Part 1 was designed to build the API call, validate the user's inputs, and return a tidy tibble with a `"census"` class. In Part 2 we took advantage of this class setting by defining our own summarizing and plot functions, which their generic counterparts route to because of that class attribute. Both methods account for the survey weight `PWGTP`, so the results reflect the population rather than the sampled rows. We also went through examples along the way to demonstrate code functionality. For instance, we observed that North Carolina's age distribution was very stable from 2021 to 2023, and that females living there tend to be a few years older than males. The project could be improved by allowing the user to choose from a list of plot types that they want to visualize their selected data in.

