---
title: "ST 558 Project 1"
author: Ryan Clark
format: html
---

## Intro

In this project we will query the Public Use Microdata Sample (PUMS) Census API. This API provides psuedo person-level data from the American Community Survey. We will focus on retrieving and summarizing data from four survey years: 2021, 2022, 2023 and 2024. 

## Part 1: Data Processing

Every year surveyed by the Census API has its own `variables.json` endpoint that lays out each variable used and their respective definitions. While there is plenty of overlap year-to-year, important differences exist that we must account for when parsing variable values from the JSON files. In this first step, we build a function that retrieves these files, extracts only the variables relevant to our analysis, and stores the result so we can reference it later.

```{r}
# Function to `GET` a PUMS `variables.json` endpoint 
get_pums_variables <- function(url, vars_to_keep) {
  var_info <- httr::GET(url)  # Contact API 
  parsed_vars <- jsonlite::fromJSON(rawToChar(var_info$content))  # Parse the raw JSON
  parsed_vars$variables[vars_to_keep]
}

# Variables found in all four years
common_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
                  "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
                  "REGION", "DIVISION")

# ST/STATE variable changes by year, so append to each year's variable list
vars_2021 <- c(common_vars, "ST")
vars_2022 <- c(common_vars, "ST")
vars_2023 <- c(common_vars, "STATE")
vars_2024 <- c(common_vars, "STATE")

# URLs
url_2021 <- "https://api.census.gov/data/2021/acs/acs1/pums/variables.json"
url_2022 <- "https://api.census.gov/data/2022/acs/acs1/pums/variables.json"
url_2023 <- "https://api.census.gov/data/2023/acs/acs1/pums/variables.json"
url_2024 <- "https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

# Build one list per year, then merge into a single list
pums_variables <- list(
  "2021" = get_pums_variables(url_2021, vars_2021),
  "2022" = get_pums_variables(url_2022, vars_2022),
  "2023" = get_pums_variables(url_2023, vars_2023),
  "2024" = get_pums_variables(url_2024, vars_2024)
)

# Save list as an .rds in the `data` folder
saveRDS(pums_variables, file = "data/pums_variables.rds")
```


```{r}
#| include: false
# API key
census_key <- Sys.getenv("CENSUS_API_KEY")
```

(placeholder text)... The variable dictionary in variables.json from the RDS has fixed-width of three characters but the API doesn't always return this so we must insert leading zeros where applicable to use the dictionary. The exception is the "N/A" code which is simply "0".

```{r}
# Helper function: Handling the time-coded variables (JWAP/JWDP)
parse_pums_time <- function(codes, code_labels) {
  padded_codes <- ifelse(codes == "0", "0",  # If 0, keep (it's the N/A code)
                         formatC(as.integer(codes), width = 3, flag = "0"))  # Otherwise convert int and insert 0s to get fixed-width

  labels <- code_labels[padded_codes]  # Match padded code to dictionary label

  # Logical vector using language ("to") to identify time-range observations
  has_range <- !is.na(labels) & grepl(" to ", labels, fixed = TRUE)

  # Isolate times
  starts <- trimws(gsub("\\.", "", sub(" to .*$", "", labels[has_range])))  # Before "to"
  ends   <- trimws(gsub("\\.", "", sub("^.*? to ", "", labels[has_range])))  # After "to"

  # Convert string times to datetime objects
  start_times <- lubridate::parse_date_time(starts, orders = "IMp", tz = "UTC")  
  end_times   <- lubridate::parse_date_time(ends,   orders = "IMp", tz = "UTC")

  # Compute midpoint time
  mid_times <- start_times + (end_times - start_times) / 2

  # Initialize empty datetime vector
  result <- as.POSIXct(rep(NA, length(labels)), tz = "UTC")
  # Fill it in with midpoint values
  result[has_range] <- mid_times
  # Return vector
  result
}
```

(placeholder text)... We will use this function to grab the raw codes of these categorical variables and replace them with their respective labels. And in the process, we'll convert this character vector into a factor while maintaining the coded order. This function will be called for every categorical variable we encounter.

```{r}
# Helper function: Replaces categorical raw character codes with corresponding label
pums_to_factor <- function(codes, code_labels) {
  ordered_codes  <- names(code_labels)[order(as.numeric(names(code_labels)))]  # Numerical ordering of codes
  ordered_labels <- unname(code_labels[ordered_codes])  # Labels ordered by their codes
  
  # Return categorical factor
  factor(unname(code_labels[codes]), levels = ordered_labels)
}
```

(placeholder text)

```{r}
# Function to query the PUMS API (year, variables, and geography) then return a parsed tibble. 
query_pums_data <- function(year = 2024,
                            numeric_vars = "AGEP",
                            categorical_vars = "SEX",
                            geography = "State",
                            geo_subset = NULL) {

  # Validate year
  valid_years <- c(2021, 2022, 2023, 2024)
  if (length(year) != 1 || !(year %in% valid_years)) {  # Not exactly 1 year OR not in list
    stop("Error: `year` must be 2021, 2022, 2023, or 2024.")
  }

  # Validate numeric variables
  valid_numeric <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if (length(numeric_vars) < 1) {  # Empty input
    stop("Error: At least one numeric variable (other than PWGTP) must be specified.")
  }
  if (!all(numeric_vars %in% valid_numeric)) {  # Checking if any vars are invalid
    stop("Error: Invalid numeric variable(s) provided. Valid options are: AGEP, GASP, GRPIP, JWAP, JWDP, and JWMNP.")
  }
  # PWGTP always returned
  numeric_vars <- unique(c(numeric_vars, "PWGTP"))

  # Validate categorical variables
  valid_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if (length(categorical_vars) < 1) {  # Empty input
    stop("Error: At least one categorical variable must be specified.")
  }
  if (!all(categorical_vars %in% valid_categorical)) {  # Checking if any vars are invalid
    stop("Error: Invalid categorical variable(s) provided. Valide options are: FER, HHL, SCH, SCHL, and SEX.")
  }

  # Validate geography level
  valid_geo <- c("Region", "Division", "State")
  if (length(geography) != 1 || !(geography %in% valid_geo)) {  # Not exactly 1 level OR not in list
    stop("Error: `geography` must be exactly Region, Division, or State.")
  }

  # Geo_subset default
  default_geo_subset <- list(Region = "3", Division = "5", State = "37")
  if (is.null(geo_subset)) {  # Empty input
    geo_subset <- default_geo_subset[[geography]] # Picks the correct default value based on the chosen geo
  }

  geo_for_name <- tolower(geography)  # Convert to lowercase for URL syntax
  geo_var_name <- switch(geography,  # Col name to return
    "Region" = "REGION",
    "Division" = "DIVISION",
    "State" = if (year %in% c(2021, 2022)) "ST" else "STATE"
  )

  # String for the URL `for=`
  geo_value <- if (identical(geo_subset, "all")) "*"  # All values (API syntax)
               else geo_subset  # User's specific choice

  # Variables list
  get_vars <- unique(c(numeric_vars, categorical_vars, geo_var_name))

  # URL
  base_url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums")
  url <- paste0(
    base_url,
    "?get=", paste(get_vars, collapse = ","),
    "&for=", geo_for_name, ":", geo_value,
    "&key=", census_key
  )

  # Contact API endpoint
  resp <- httr::GET(url)
  raw_data <- jsonlite::fromJSON(rawToChar(resp$content))

  # Initialize tibble
  pums_tibble <- raw_data |>
  tibble::as_tibble(.name_repair = "minimal") |>  # Prevent auto-naming
  janitor::row_to_names(row_number = 1)  # First row is col headers, removed from data

  # Load RDS and convert year to char
  variable_info <- readRDS("data/pums_variables.rds")
  year_info <- variable_info[[as.character(year)]]

  # Convert numeric variables to num or POSIXct
  time_vars <- c("JWAP", "JWDP")  # Arrival/departure times
  for (var in numeric_vars) {
    if (var %in% time_vars) {  # is the variable time-coded?
      code_labels <- unlist(year_info[[var]]$values$item)  # Table of codes and labels
      pums_tibble[[var]] <- parse_pums_time(pums_tibble[[var]], code_labels)  # Helper function
    } else {
      pums_tibble[[var]] <- as.numeric(pums_tibble[[var]]) # Otherwise numeric
    }
  }

  # Convert categorical variables to labeled factors
  for (var in categorical_vars) {
    code_labels <- unlist(year_info[[var]]$values$item)  # Table of codes and labels
    pums_tibble[[var]] <- pums_to_factor(pums_tibble[[var]], code_labels) # Helper function
  }

  # Adding the "census" class to the tibble
  class(pums_tibble) <- c("census", class(pums_tibble))
  
  # Return tibble
  pums_tibble
}
```

(placeholder text)... The function will take a vector of years, not simply one year... The tibble returned will have a "year" column identifying the source year.

```{r}
# Function to query multiple data years then return a tibble.
query_pums_multi_year <- function(years,  # Vector of years (unlike query_pums_data)
                                  numeric_vars = "AGEP",
                                  categorical_vars = "SEX",
                                  geography = "State",
                                  geo_subset = NULL) {

  # Loop over each element of `years`, applies function to assemble list of tibbles
  results <- lapply(years, function(yr) {
    result <- query_pums_data(  # A year's raw tibble
      year = yr,
      numeric_vars = numeric_vars,
      categorical_vars = categorical_vars,
      geography = geography,
      geo_subset = geo_subset
    )

    # Changing state variable names to all be "STATE" (2021/2022)
    if ("ST" %in% names(result)) {
      names(result)[names(result) == "ST"] <- "STATE"
    }

    # Record which survey year this data came from
    result$year <- yr

    # Strip the "census" class before combining (for dplyr)
    class(result) <- class(result)[class(result) != "census"]
    
    # Return a year's tibble
    result
  })
  
  # Combine (vertically)
  combined <- dplyr::bind_rows(results)

  # Adding back the "census" class to the combined result
  class(combined) <- c("census", class(combined))

  # Return tibble
  combined
}
```








